Interactively Visualizing Clustering in Multivariate Data Using lionfish

### 2024-09-27 manuscript.tex
changed variable to feature and adjusted the metric descriptions

A difference between clustering analysis and partitioning is typically the nature of the data. With cluster analysis, we usually envision data that contains separated clusters, and a successful clustering result is one that divides the data based on these gaps. With partitioning, it is usual that there are no gaps in the data, but it is still useful to partition the data. Figure \ref{kmeans-partition} illustrates how the $k$-means algorithm would partition a 2D data set into four groups depending on the correlation between the two variables.  When the correlation is high, the partitioning will be along the combination of variables that produces the highest variance. With lower correlation, it will segment the bottom and top, and divide the middle into two parts in the opposite direction. When there is no association, the partitioning is radial like a windmill.

@@

One can see that the data is perfectly divided into four parts. However, if one were to plot the two variables individually, this would not be obvious. Figure \ref{kmeans-histogram} shows histograms of the two variables, V1, V2, with the colour matching the four partitions. From the histograms, we can see some differences in the partitions for the four different association structures, but they are all overlapping. The distinct border between the partitions can only be seen from the scatterplots of both variables.

\begin{figure*}[ht]
\centerline{\includegraphics[width=1\textwidth]{images/intro2.png}}
\caption{The four $k$-means partitions plotted as histograms of the individual variables. While differences between groups can be seen, the clear separation cannot. This is important for understanding why high-dimensional visualisation methods are useful for summarising partitioning results.}
\label{kmeans-histogram}
\end{figure*}

When there are more than two variables, histograms of the individual variables are commonly used to display the partitioning results. This means that the analyst likely cannot understand how the partitioning divides the data. All that they can observe is roughly how the individual variables relate to the partition, which is useful but inadequate. Here is where using tour methods to view high dimensions can be helpful. A tour (\cite{Asimov1985-xr}, and see \cite{lee2022}) can be used to show scatterplots of combinations of variables and thus provide views like that in Figure \ref{kmeans-partition} where distinct differences between partitions can be observed. High dimensions are still tricky, and a combination of animations of the linear combinations, and interactive control \citep{cook_manual_1997, laa2023new} over the combinations is important. A scatterplot of a combination of variables can be considered to be a projection of the data, and thus like a shadow of a 3D object, some aspects of the data (object) can be obscured. Using slices of the projected data \citep{Laa2020} can be a useful addition to projections. This paper illustrates how to do this to better understand partitioning results for multivariate data.

@@

To integrate user interactions with the capabilities of the tourr package an active communication with the interface is required. For example, we may wish to explore the local neighbourhood of a projection selected by the user with a local tour animation provided by tourr, or we may want to optimize a guided tour path using groups identified via brushing.
Our solution is using Python for high-performance interactivity, through the packages \texttt{TKinter} \citep{lundh1999introduction}, \texttt{CustomTKinter} \citep{schimansky24}, and \texttt{matplotlib} \citep{Hunter:2007}, with integration to the tourr package \citep{tourr} via \texttt{reticulate} \citep{reticulate}, a framework that facilitates seamless interoperability between Python and R.
\textcolor{red}{also rpy2?}

@@

The interface was implemented in \texttt{lionfish} and offers a variety of linked interactive plot types, providing users with the flexibility to visualize their data from multiple perspectives. The ability to navigate through various projections of the displayed tours directly within a graphical user interface (GUI) enables users to explore different aspects of the dataset. Furthermore, users can initiate new tours directly from the interface. The GUI also supports interactive variable selection, allowing users to specify which subset of variables should be visualized in the plots. Once users have identified interesting views or settings, \texttt{lionfish} allows them to save the displayed projections, subsets, and plots. This functionality ensures that analysis states can be preserved for further examination or reporting, making the package particularly useful for iterative analysis where findings may need to be revisited or shared with collaborators.

@@

\subsection{Variable selection}

A first step may look into what variables are important for the cluster solution. The interface provides different capabilities that support variable selection. First, a 2D tour view can be used to understand the sensitivity of the grouping to individual variables, in particular when using the manual tour to remove one variable at a time. This is in particular useful when the starting view was obtained through a projection pursuit guided tour~\citep{CBCH94} that optimized for the separation between the labeled groups~\citep{lckl2005}.

In our applications we are considering market segmentation analysis on binary variables. For this setting we have developed a heatmap display that we have found to be particularly useful in variable selection. The idea is to understand differences in the binary ratings between clusters, where we use different normalizations to better explore the data. For \(c_{ij}\) the count (number of true values in the binary variable) in the \(i\)-th cluster and \(j\)-th variable, we define

\[
\text{Overall count fraction}\ f_{ij} = \frac{c_{ij}}{n}
\]
with normalization to the total number of observations $n$ for a general overview,

\[
\text{Cluster count fraction}\  f_{ij} = \frac{c_{ij}}{n_{i}}
\]
showing the counts relative to the cluster's size \(n_i\), to see the fraction of observations in cluster $i$ for which variable $j$ was true, and

\[
\text{Variable count fraction}\ f_{ij} = \frac{c_{ij}}{c_{j}}
\]
to see how the true values for variable $j$ are distributed across the different clusters relative, with normalization to the overall count of that feature \(c_j = \sum_i c_{ij}\).

@@

\begin{itemize}
    \item A \texttt{.png} file containing the currently displayed graphics,
    \item \texttt{.csv} files that capture the feature and subset selection as well as projections of the tours displayed at the time of the snapshot,
    \item two \texttt{.pkl} files that contain state variables of the GUI, allowing for complete recovery of the snapshot.
\end{itemize}

@@

The Austrian Vacation Activities dataset comprises responses from 2,961 adult tourists who spent their holiday in Austria during the 1997/98 season. Participants were asked to evaluate the importance of 27 different activities during their vacation. The survey categorized responses based on four levels of importance: ``totally important'', ``mostly important'', ``a bit important'', and ``not important''. For analysis, the responses were binarized: a value of 1 was assigned if the activity was rated as ``totally important'', and a value of 0 if any of the other categories were selected. The survey was conducted by the Europäisches Tourismus Institut GmbH at the University of Trier and focused exclusively on tourists who did not stay in the country's capital cities. Figure \ref{winter-gt} shows three projections from a grand tour of this data: when mostly two variables contribute to a projection the data will look apparently clustered, due to the binary values, but when more variables contribute to the projection it looks more continuous.

\begin{figure*}[h]
\centerline{\includegraphics[width=0.3\textwidth]{images/winter_proj1.pdf}\includegraphics[width=0.3\textwidth]{images/winter_proj2.pdf}\includegraphics[width=0.3\textwidth]{images/winter_proj3.pdf}}
\caption{Three 2D projections from a grand tour on the Austrian Vacation Activities data. Because the 27 variables are binary, you'll see some apparent clusters (left) but with this many binary variables the data mostly looks continuous.}
\label{winter-gt}
\end{figure*}

>>>>

A difference between clustering analysis and partitioning is typically the nature of the data. With cluster analysis, we usually envision data that contains separated clusters, and a successful clustering result is one that divides the data based on these gaps. With partitioning, it is usual that there are no gaps in the data, but it is still useful to partition the data. Figure \ref{kmeans-partition} illustrates how the $k$-means algorithm would partition a 2D data set into four groups depending on the correlation between the two features. When the correlation is high, the partitioning will be along the combination of features that produces the highest variance. With lower correlation, it will segment the bottom and top, and divide the middle into two parts in the opposite direction. When there is no association, the partitioning is radial like a windmill.

@@

One can see that the data is perfectly divided into four parts. However, if one were to plot the two features individually, this would not be obvious. Figure \ref{kmeans-histogram} shows histograms of the two features, V1, V2, with the colour matching the four partitions. From the histograms, we can see some differences in the partitions for the four different association structures, but they are all overlapping. The distinct border between the partitions can only be seen from the scatterplots of both features.

\begin{figure*}[ht]
\centerline{\includegraphics[width=1\textwidth]{images/intro2.png}}
\caption{The four $k$-means partitions plotted as histograms of the individual features. While differences between groups can be seen, the clear separation cannot. This is important for understanding why high-dimensional visualisation methods are useful for summarising partitioning results.}
\label{kmeans-histogram}
\end{figure*}

When there are more than two features, histograms of the individual features are commonly used to display the partitioning results. This means that the analyst likely cannot understand how the partitioning divides the data. All that they can observe is roughly how the individual features relate to the partition, which is useful but inadequate. Here is where using tour methods to view high dimensions can be helpful. A tour (\cite{Asimov1985-xr}, and see \cite{lee2022}) can be used to show scatterplots of combinations of features and thus provide views like that in Figure \ref{kmeans-partition} where distinct differences between partitions can be observed. High dimensions are still tricky, and a combination of animations of the linear combinations, and interactive control \citep{cook_manual_1997, laa2023new} over the combinations is important. A scatterplot of a combination of features can be considered to be a projection of the data, and thus like a shadow of a 3D object, some aspects of the data (object) can be obscured. Using slices of the projected data \citep{Laa2020} can be a useful addition to projections. This paper illustrates how to do this to better understand partitioning results for multivariate data.

@@

To integrate user interactions with the capabilities of the tourr package an active communication with the interface is required. For example, we may wish to explore the local neighbourhood of a projection selected by the user with a local tour animation provided by tourr, or we may want to optimize a guided tour path using groups identified via brushing.
Our solution is using Python for high-performance interactivity, through the packages \texttt{TKinter} \citep{lundh1999introduction}, \texttt{CustomTKinter} \citep{schimansky24}, and \texttt{matplotlib} \citep{Hunter:2007}, with integration to the tourr package \citep{tourr} via \texttt{reticulate} \citep{reticulate}, a framework that facilitates seamless interoperability between Python and R.

@@

The interface was implemented in \texttt{lionfish} and offers a variety of linked interactive plot types, providing users with the flexibility to visualize their data from multiple perspectives. The ability to navigate through various projections of the displayed tours directly within a graphical user interface (GUI) enables users to explore different aspects of the dataset. Furthermore, users can initiate new tours directly from the interface. The GUI also supports interactive feature selection, allowing users to specify which subset of features should be visualized in the plots. Once users have identified interesting views or settings, \texttt{lionfish} allows them to save the displayed projections, subsets, and plots. This functionality ensures that analysis states can be preserved for further examination or reporting, making the package particularly useful for iterative analysis where findings may need to be revisited or shared with collaborators.

@@

\subsection{Feature selection based on fractional analysis}

A first step may look into what features are important for the cluster solution. The interface provides different capabilities that support feature selection. First, a 2D tour view can be used to understand the sensitivity of the grouping to individual features, in particular when using the manual tour to remove one feature at a time. This is in particular useful when the starting view was obtained through a projection pursuit guided tour~\citep{CBCH94} that optimized for the separation between the labeled groups~\citep{lckl2005}.

In our applications we are considering market segmentation analysis on binary features. For this setting we have developed a heatmap display that we have found to be particularly useful in feature selection. The idea is to understand differences in the binary ratings between clusters, where we use different normalizations to better explore the data. For \(c_{ij}\) the count (number of true values in the binary feature) in the \(i\)-th cluster and \(j\)-th feature, we define

\[
\text{Total fraction}\ f_{ij}^{tot} = \frac{c_{ij}}{n}
\]
with normalization to the total number of observations $n$ for a general overview,

\[
\text{Intra cluster fraction}\  f_{ij}^{clust} = \frac{c_{ij}}{n_{i}}
\]
showing the counts relative to the cluster's size \(n_i\), to see the fraction of observations in cluster $i$ for which feature $j$ was true, and

\[
\text{Intra feature fraction}\ f_{ij}^{feat} = \frac{c_{ij}}{c_{j}}
\]
to see how the true values for feature $j$ are distributed across the different clusters relative, with normalization to the overall count of that feature \(c_j = \sum_i c_{ij}\).

@@

\begin{itemize}
    \item A \texttt{.png} file containing the currently displayed graphics,
    \item \texttt{.csv} files that capture the feature and subset selection as well as projections of the tours displayed at the time of the snapshot,
    \item two \texttt{.pkl} files that contain state features of the GUI, allowing for complete recovery of the snapshot.
\end{itemize}

@@

The Austrian Vacation Activities dataset comprises responses from 2,961 adult tourists who spent their holiday in Austria during the 1997/98 season. Participants were asked to evaluate the importance of 27 different activities during their vacation. The survey categorized responses based on four levels of importance: ``totally important'', ``mostly important'', ``a bit important'', and ``not important''. For analysis, the responses were binarized: a value of 1 was assigned if the activity was rated as ``totally important'', and a value of 0 if any of the other categories were selected. The survey was conducted by the Europäisches Tourismus Institut GmbH at the University of Trier and focused exclusively on tourists who did not stay in the country's capital cities. Figure \ref{winter-gt} shows three projections from a grand tour of this data: when mostly two features contribute to a projection the data will look apparently clustered, due to the binary values, but when more features contribute to the projection it looks more continuous.

\begin{figure*}[h]
\centerline{\includegraphics[width=0.3\textwidth]{images/winter_proj1.pdf}\includegraphics[width=0.3\textwidth]{images/winter_proj2.pdf}\includegraphics[width=0.3\textwidth]{images/winter_proj3.pdf}}
\caption{Three 2D projections from a grand tour on the Austrian Vacation Activities data. Because the 27 features are binary, you'll see some apparent clusters (left) but with this many binary features the data mostly looks continuous.}
\label{winter-gt}
\end{figure*}
